Accelerated Inference for Partially Observed Stochastic Processes with Automatic Differentiation

### 2024-07-10 paper/appendix/biasvar.tex
Updates from Overleaf

\begin{proof}
When $\alpha=0$, we have:
\begin{align}
    \hat{\lik}^0(\theta) := \prod_{n=1}^N L_n^{A, \theta, 0} &= \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta,0}}{\sum_{j=1}^J w_{n,j}^{P,\theta,0}} 
    \\&
    = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J s_{n,j} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J \frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \theta})}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi})}.
\end{align}
Similarly to the proof of Lemma \ref{lem:mop-0-formula}, we define $s_{n,j}=\frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \theta})}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi})}$, and from the exact same arguments conclude that its gradient when $\theta=\phi$ is therefore 
\begin{align}
    \nabla_\theta \hat{\ell}^0(\theta) &:= \sum_{n=1}^N \nabla_\theta \log\left(L_n^\phi \frac{1}{J} \sum_{j=1}^J s_{n,j}\right) \\
    &= \sum_{n=1}^N \frac{\nabla_\theta \left(L_n^\phi \frac{1}{J} \sum_{j=1}^J s_{n,j}\right)}{\left(L_n^\phi \frac{1}{J} \sum_{j=1}^J s_{n,j}\right)} \\
    &= \sum_{n=1}^N \frac{\sum_{j=1}^J \nabla_\theta s_{n,j}}{\sum_{j=1}^J s_{n,j}} \\
    &= \sum_{n=1}^N \frac{1}{J} \sum_{j=1}^J \frac{\nabla_\theta f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)}{f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \phi}; \phi)} \\
    &= \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right),
\end{align}
where we use the log-derivative trick, observe that $\sum_{j=1}^J s_{n,j} = J$ when $\theta=\phi$, and use the log-derivative trick where $\theta=\phi$ again. We do this to establish that the gradient of the log-likelihood estimate is given by the sum of terms over all $N$ and $J$:
\begin{equation}
\nabla_\theta \hat{\ell}^0(\theta) = \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right).
\end{equation}
Therefore, for a given $\theta_i$ in the parameter vector $\theta$,
\begin{align}
    \Var\left(\frac{\partial}{\partial_{\theta_i}} \hat\ell^0(\theta)\right) &= \frac{1}{J^2}\Var\left(\sum_{n=1}^N\sum_{j=1}^{J}\frac{\partial}{\partial_{\theta_i}} \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
    & \hspace{-20mm}= \frac{1}{J^2}\sum_{n=1}^N\Var\left(\sum_{j=1}^{J}\frac{\partial}{\partial_{\theta_i}} \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
    &\hspace{-20mm}+ 2\sum_{m<n}\Cov\left(\frac{1}{J}\sum_{j=1}^{J}\frac{\partial}{\partial_{\theta_i}} \log\left(f_{Y_m|X_{m}}(y_m^*|x_{m,j}^{F, \theta}; \theta)\right), \frac{1}{J}\sum_{j=1}^{J}\frac{\partial}{\partial_{\theta_i}} \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
    &\hspace{-20mm}= \sum_{n=1}^N \Var\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right) + 2\sum_{m<n} \Cov\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_m^0(\theta),\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right).
\end{align}
Here, we use Assumptions \ref{assump:bounded-process} and \ref{assump:bounded-measurement} that ensure strong mixing. We know from Theorem 3 of \cite{karjalainen23} that when $\textbf{M}_{n,n+k}$ is the $k$-step Markov operator from timestep $n$ and $\beta_{\text{TV}}(M) = \sup _{x, y \in E}\|M(x, \cdot)-M(y, \cdot)\|_{\mathrm{TV}}=\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu M-\nu M\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}}$ is the Dobrushin contraction coefficient of a Markov operator, 
\begin{equation}
\beta_{\text{TV}}(\textbf{M}_{n,n+k}) \leq (1-\epsilon)^{\floor{k/(c\log(J))}},
\end{equation}
i.e. the mixing time of the particle filter is $O(\log(J))$, where $\epsilon$ and $c$ depend on $\bar{M}, \underbar{M}, \bar{G}, \underbar{G}$ in \ref{assump:bounded-process} and \ref{assump:bounded-measurement}.

@@

By Lemma \ref{lem:dobrushin-implies-alpha-mixing}, the particle filter itself is strong mixing, with $\alpha$-mixing coefficients $\alpha(k) \leq (1-\epsilon)^{\floor{k/(c\log(J))}}$. Therefore, functions of particles are strongly mixing as well, with $\alpha$-mixing coefficients bounded by the original (to see this, observe that the $\sigma$-algebra of the functionals is contained within the original $\sigma$-algebra). Therefore, by Davydov's inequality, noting that $\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\leq G'(\theta)$ by Assumption \ref{assump:bounded-measurement}, and without loss of generality labeling $m$ and $n$ such that $\E\big[(\frac{\partial}{\partial_{\theta_i}}\hat\ell_m^0(\theta))^4\big]^{1/4}\leq\E\big[(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta))^4\big]^{1/4}$, we see that
\begin{align}
    \Cov\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_m^0(\theta), \frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right) 
    &\leq \alpha(n-m)^{1/2} \, \E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_m^0(\theta)\right)^4\right]^{1/4}
    \,
    \E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right)^4\right]^{1/4}\\
    &\leq \alpha(n-m)^{1/2}\, \E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right)^4\right]^{1/2}.
\end{align}
To bound this, we use the fact that 
\begin{equation}
\E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right)^4\right] = \E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)-\E\left[\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right]\right)^4\right]+\E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right)^2\right]^2
\end{equation}
alongside Lemma 2 of \cite{karjalainen23}, which shows that
\begin{eqnarray}
\E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)-\E\left[\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right]\right)^4\right] &\lesssim& \frac{G'(\theta)^4}{J^2}, 
\\
\E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)-\E\left[\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right]\right)^2\right] &\lesssim& \frac{G'(\theta)^2}{J}.
\end{eqnarray}
It follows that 
\begin{equation}
\E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right)^4\right]^{1/2} \lesssim  \sqrt{\frac{G'(\theta)^4}{J^2}+\left(\frac{G'(\theta)^2}{J}\right)^2} = \frac{G'(\theta)^2}{J},
\end{equation}
and we conclude that 
\begin{equation}
\Cov\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_m^0(\theta), \frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right) \leq (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c\log(J)}}}\frac{G'(\theta)^2}{J}.
\end{equation}
Putting it all together, we see that
\begin{align}
    \Var\left(\frac{\partial}{\partial_{\theta_i}} \hat\ell(\theta)\right) &= \sum_{n=1}^N \Var\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right) + 2\sum_{m<n} \Cov\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_m^0(\theta),\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right)\\
    &\leq \frac{NG'(\theta)^2}{J} + 2\sum_{n=1}^{N-1} (N-n)(1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c\log(J)}}}\frac{G'(\theta)^2}{J} \\
    &\leq \frac{G'(\theta)^2}{J} \left(N + 2\sum_{n=1}^{N-1} (N-n) (1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c\log(J)}}}\right) \\
    &\lesssim \frac{G'(\theta)^2N}{J}.
\end{align}
It then follows that as $\theta \in \Theta \subseteq \R^p$, 
\begin{equation}
    \Var\big(\nabla_{\theta} \hat\ell(\theta)\big) \lesssim \frac{Np \, G'(\theta)^2}{J}.
\end{equation}
\end{proof}

@@

Although the variance of MOP-$0$ is desirable, it unfortunately has large uncontrollable asymptotic bias. Error bounds for the bias and MSE of MOP-$0$ \cite{naesseth18} are also fortunately special cases (where $k=1$, $\alpha=0$) of the bias and MSE analysis we conduct for MOP-$\alpha$. We will later see that these showcase the large uncontrollable asymptotic bias of MOP-$0$.

@@

We are now in a position to tackle the MOP-$\alpha$ variance bound. Here, we analyze the case for $\alpha \in (0,1)$. The proof strategy is as follows. Instead of analyzing the variance of the MOP-$\alpha$ gradient estimate proper, we will analyze the variance of a modified estimator where we truncate the weights at $k$ timesteps, so this estimator only looks at most $k$ timesteps back while accumulating the importance weights. This enables us to establish strong mixing for the truncated estimator, in order to get a bound on the variance that is $O(NJ^{-1})$. We emphasize that this estimator is only a theoretical construct -- we do not actually use this estimator.

@@

\begin{thm}[Variance of MOP-$\alpha$ Gradient Estimate]
    When $\alpha \in (0,1)$, the variance of the gradient estimate from MOP-$\alpha$ is
    \begin{equation}\Var\big(\nabla_\theta \hat\ell(\theta)\big) \lesssim \min_{k\leq N} \left(\frac{k^2G'(\theta)^2Np}{(1-\alpha)^2J} + \frac{\alpha^{k}}{1-\alpha}Np \, G'(\theta)^2 \right).
    \end{equation}
\end{thm}
\begin{proof}

Using the log-derivative trick and that $w_{n,j}^{F,\theta} = w_{n,j}^{F,\theta,k} = 1$ when $\theta=\phi$,
\begin{align}
    &\left\lVert\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta}\right)-\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,k}\right)\right\rVert_2^2\\
    &= \left\lVert\frac{\nabla_\theta\sum_{j=1}^J w_{n,j}^{F,\theta}}{{\sum_{j=1}^J w_{n,j}^{F,\theta}}}-\frac{\nabla_\theta\sum_{j=1}^J w_{n,j}^{F,\theta,k}}{{\sum_{j=1}^J w_{n,j}^{F,\theta,k}}}\right\rVert_2^2 \\
    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta}-\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta,k}\right\rVert_2^2 \\
    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \frac{\nabla_\theta w_{n,j}^{F,\theta}}{w_{n,j}^{F,\theta}}-\frac{1}{J}\sum_{j=1}^J \frac{\nabla_\theta w_{n,j}^{F,\theta,k}}{w_{n,j}^{F,\theta,k}}\right\rVert_2^2\\
    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta}\right)-\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta,k}\right)\right\rVert_2^2.
\end{align}

@@

    &  \left\lVert\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta}\right)-\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,k}\right)\right\rVert_2^2\\
    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta}\right)-\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta,k}\right)\right\rVert_2^2 \\
    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\log\left(w_{n,j}^{F,\theta}\right)-\log\left(w_{n,j}^{F,\theta,k}\right)\right)\right\rVert_2^2\\

@@

We can then bound $\|\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta)\|$ as follows:

@@

    &\left\lVert\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta) \right\rVert_2^2

@@

    &= \left\lVert\sum_{n=1}^N \nabla_\theta \Bigg(\!\log\!\left(\sum_{j=1}^Jw_{n,j}^{F,\theta}\right) \!-\! \log\!\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,k}\right)
    \!-\! \log\!\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta}\right) \!+\! \log \! \left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,k}\right)\!\Bigg)\right\rVert_2^2 

@@

    &\leq \sum_{n=1}^N \left\lVert\nabla_\theta \left(\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,k}\right)\right)\right\lVert_{\infty}

@@

    & \qquad\qquad +\sum_{n=1}^N \left\lVert\nabla_\theta \left(\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,k}\right)\right)\right\rVert_2^2,

@@

\begin{equation}\left\lVert\nabla_\theta\hat\ell(\theta) - \hat s_k(\theta) \right\rVert_2^2 \leq  \frac{2\alpha^k}{1-\alpha}Np \, G'(\theta)^2,\end{equation}

@@

Now, we bound the variance of $\hat s_k(\theta)$. 

@@

We can then decompose this expression with the log-derivative trick while noting that $w_{n,j}^{F,\theta, k} = 1$ whenever $\theta=\phi$, to see that

>>>>

\begin{proof}
When $\alpha=0$, we have:
\begin{align}
    \hat{\lik}^0(\theta) := \prod_{n=1}^N L_n^{A, \theta, 0} &= \prod_{n=1}^N L_n^\phi \cdot \frac{\sum_{j=1}^J w_{n,j}^{F,\theta,0}}{\sum_{j=1}^J w_{n,j}^{P,\theta,0}} 
    \\&
    = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J s_{n,j} = \prod_{n=1}^N L_n^\phi \cdot \frac{1}{J}\sum_{j=1}^J \frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \theta})}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi})}.
\end{align}
Similarly to the proof of Lemma \ref{lem:mop-0-formula}, we define 
\begin{equation}
s_{n,j}=\frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \theta})}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi})},
\end{equation}
and from the exact same arguments conclude that its gradient when $\theta=\phi$ is therefore 
\begin{align}
    \nabla_\theta \hat{\ell}^0(\theta) &:= \sum_{n=1}^N \nabla_\theta \log\left(L_n^\phi \frac{1}{J} \sum_{j=1}^J s_{n,j}\right) \\
    &= \sum_{n=1}^N \frac{\nabla_\theta \left(L_n^\phi \frac{1}{J} \sum_{j=1}^J s_{n,j}\right)}{\left(L_n^\phi \frac{1}{J} \sum_{j=1}^J s_{n,j}\right)} \\
    &= \sum_{n=1}^N \frac{\sum_{j=1}^J \nabla_\theta s_{n,j}}{\sum_{j=1}^J s_{n,j}} \\
    &= \sum_{n=1}^N \frac{1}{J} \sum_{j=1}^J \frac{\nabla_\theta f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)}{f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \phi}; \phi)} \\
    &= \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right),
\end{align}
where we use the derivative of the logarithm, observe that $\sum_{j=1}^J s_{n,j} = J$ when $\theta=\phi$, and use the derivative of the logarithm where $\theta=\phi$ again. We do this to establish that the gradient of the log-likelihood estimate is given by the sum of terms over all $N$ and $J$:
\begin{equation}
\nabla_\theta \hat{\ell}^0(\theta) = \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right).
\end{equation}
Therefore, for a given $\theta_i$ in the parameter vector $\theta$,
\begin{align}
    \Var\left(\frac{\partial}{\partial_{\theta_i}} \hat\ell^0(\theta)\right) &= \frac{1}{J^2}\Var\left(\sum_{n=1}^N\sum_{j=1}^{J}\frac{\partial}{\partial_{\theta_i}} \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
    & \hspace{-20mm}= \frac{1}{J^2}\sum_{n=1}^N\Var\left(\sum_{j=1}^{J}\frac{\partial}{\partial_{\theta_i}} \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
    &\hspace{-20mm}+ 2\sum_{m<n}\Cov\left(\frac{1}{J}\sum_{j=1}^{J}\frac{\partial}{\partial_{\theta_i}} \log\left(f_{Y_m|X_{m}}(y_m^*|x_{m,j}^{F, \theta}; \theta)\right), \frac{1}{J}\sum_{j=1}^{J}\frac{\partial}{\partial_{\theta_i}} \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right)\right) \\
    &\hspace{-20mm}= \sum_{n=1}^N \Var\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right) + 2\sum_{m<n} \Cov\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_m^0(\theta),\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right).
\end{align}
Here, we use Assumptions \ref{assump:bounded-process} and \ref{assump:bounded-measurement} that ensure strong mixing. We know from Theorem 3 of \cite{karjalainen23} that when $\textbf{M}_{n,n+k}$ is the $k$-step Markov operator from timestep $n$ and $\beta_{\text{TV}}(M) = \sup _{x, y \in E}\|M(x, \cdot)-M(y, \cdot)\|_{\mathrm{TV}}=\sup _{\mu, \nu \in \mathcal{P}, \mu \neq \nu} \frac{\|\mu M-\nu M\|_{\mathrm{TV}}}{\|\mu-\nu\|_{\mathrm{TV}}}$ is the Dobrushin contraction coefficient of a Markov operator, 
\begin{equation}
\beta_{\text{TV}}(\textbf{M}_{n,n+k}) \leq (1-\epsilon)^{\floor{k/(c\log(J))}},
\end{equation}
i.e. the mixing time of the particle filter is $O(\log(J))$, where $\epsilon$ and $c$ depend on $\bar{M}, \underbar{M}, \bar{G}, \underbar{G}$ in \ref{assump:bounded-process} and \ref{assump:bounded-measurement}.

@@

By Lemma \ref{lem:dobrushin-implies-alpha-mixing}, the particle filter itself is strong mixing, with $\alpha$-mixing coefficients $a(k) \leq (1-\epsilon)^{\floor{k/(c\log(J))}}$. Therefore, functions of particles are strongly mixing as well, with $\alpha$-mixing coefficients bounded by the original (to see this, observe that the $\sigma$-algebra of the functionals is contained within the original $\sigma$-algebra). Therefore, by Davydov's inequality, noting that $\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\leq G'(\theta)$ by Assumption \ref{assump:bounded-measurement}, and without loss of generality labeling $m$ and $n$ such that $\E\big[(\frac{\partial}{\partial_{\theta_i}}\hat\ell_m^0(\theta))^4\big]^{1/4}\leq\E\big[(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta))^4\big]^{1/4}$, we see that
\begin{align}
    \Cov\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_m^0(\theta), \frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right) 
    &\leq a(n-m)^{1/2} \, \E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_m^0(\theta)\right)^4\right]^{1/4}
    \,
    \E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right)^4\right]^{1/4}\\
    &\leq a(n-m)^{1/2}\, \E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right)^4\right]^{1/2}.
\end{align}
To bound this, we use the fact that 
\begin{equation}
\E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right)^4\right] = \E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right)^4-\E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right)^2\right]^2\right]+\E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right)^2\right]^2
\end{equation}
alongside Lemma 2 of \cite{karjalainen23}, which shows that
\begin{eqnarray}
\E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right)^4-\E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right)^2\right]^2\right] &\lesssim& \frac{G'(\theta)^4}{J^2}, 
\\
\E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)-\E\left[\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right]\right)^2\right] &\lesssim& \frac{G'(\theta)^2}{J}.
\end{eqnarray}
It follows that 
\begin{equation}
\E\left[\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right)^4\right]^{1/2} \lesssim  \sqrt{\frac{G'(\theta)^4}{J^2}+\left(\frac{G'(\theta)^2}{J}\right)^2} = \frac{G'(\theta)^2}{J},
\end{equation}
and we conclude that 
\begin{equation}
\Cov\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_m^0(\theta), \frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right) \leq (1-\epsilon)^{\frac{1}{2}\floor{\frac{|n-m|}{c\log(J)}}}\frac{G'(\theta)^2}{J}.
\end{equation}
Putting it all together, we see that
\begin{align}
    \Var\left(\frac{\partial}{\partial_{\theta_i}} \hat\ell^\alpha(\theta)\right) &= \sum_{n=1}^N \Var\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right) + 2\sum_{m<n} \Cov\left(\frac{\partial}{\partial_{\theta_i}}\hat\ell_m^0(\theta),\frac{\partial}{\partial_{\theta_i}}\hat\ell_n^0(\theta)\right)\\
    &\leq \frac{NG'(\theta)^2}{J} + 2\sum_{n=1}^{N-1} (N-n)(1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c\log(J)}}}\frac{G'(\theta)^2}{J} \\
    &\leq \frac{G'(\theta)^2}{J} \left(N + 2\sum_{n=1}^{N-1} (N-n) (1-\epsilon)^{\frac{1}{2}\floor{\frac{n}{c\log(J)}}}\right) \\
    &\lesssim \frac{G'(\theta)^2N}{J}.
\end{align}
It then follows that as $\theta \in \Theta \subseteq \R^p$, 
\begin{equation}
    \Var\big(\nabla_{\theta} \hat\ell^\alpha(\theta)\big) \lesssim \frac{Np \, G'(\theta)^2}{J}.
\end{equation}
\end{proof}

@@

%Although the variance of MOP-$0$ is desirable, it unfortunately has large uncontrollable asymptotic bias. Error bounds for the bias and MSE of MOP-$0$ \cite{naesseth18} are also fortunately special cases (where $k=1$, $\alpha=0$) of the bias and MSE analysis we conduct for MOP-$\alpha$. We will later see that these showcase the large uncontrollable asymptotic bias of MOP-$0$.

@@

We are now in a position to tackle the MOP-$\alpha$ variance bound. Here, we analyze the case for $\alpha \in (0,1)$. The proof strategy is as follows. Instead of analyzing the variance of the MOP-$\alpha$ gradient estimate proper, we will analyze the variance of a modified estimator where we truncate the weights at $k$ timesteps, so this estimator only looks at most $k$ timesteps back while accumulating the importance weights. We write
\begin{equation}
    \hat s_k^\alpha(\theta) := \sum_{n=1}^N \log\left(\frac{\sum_{j=1}^J\prod_{i=n-k}^n\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}{\sum_{j=1}^J\prod_{i=n-k}^{n-1}\left(\frac{g_{i,j}^{A,F,\theta}}{g_{i,j}^{A,F,\phi}} \right)^{\alpha^{(n-i)}}}\right)
\end{equation}
for the log-likelihood estimate from the $k$-truncated estimator with discount factor $\alpha$. It is the sum of components
\begin{equation}
\hat s_{n,k}^\alpha(\theta) := \frac{1}{J}\sum_{j=1}^J\sum_{i=n-k}^n \left(\alpha^{n-i} \log\left(g_{i,j}^{A,F,\theta}\right)- \alpha^{n-i+1} \nabla_\theta \log\left(g_{i-1,j}^{A,F,\theta}\right)\right)
\end{equation} 
over timesteps $N$.

This enables us to establish strong mixing for the truncated estimator, in order to get a bound on the variance that is $O(NJ^{-1})$. We emphasize that this truncated estimator is only a theoretical construct -- we do not actually use the truncated estimator. The truncated estimator, as we note in the discussion, possesses similar theoretical guarantees as the MOP-$\alpha$ estimator, but would require the user to specify the number of timesteps to truncate at instead of a discount factor.

@@

\begin{thm}[Variance of MOP-$\alpha$ Gradient Estimate]
    When $\alpha \in (0,1)$, the variance of the gradient estimate from MOP-$\alpha$ is
    \begin{equation}\Var\big(\nabla_\theta \hat\ell^\alpha(\theta)\big) \lesssim \min_{k\leq N} \left(\frac{k^2G'(\theta)^2Np}{(1-\alpha)^2J} + \frac{\alpha^{k}}{1-\alpha}Np \, G'(\theta)^2 \right).
    \end{equation}
\end{thm}
\begin{proof}

Using the derivative of the logarithm and that $w_{n,j}^{F,\theta,\alpha} = w_{n,j}^{F,\theta,\alpha,k} = 1$ when $\theta=\phi$,
\begin{align}
    &\left\lVert\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha}\right)-\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha,k}\right)\right\rVert_2^2\\
    &= \left\lVert\frac{\nabla_\theta\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha}}{{\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha}}}-\frac{\nabla_\theta\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha,k}}{{\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha,k}}}\right\rVert_2^2 \\
    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta,\alpha}-\frac{1}{J}\sum_{j=1}^J \nabla_\theta w_{n,j}^{F,\theta,\alpha,k}\right\rVert_2^2 \\
    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \frac{\nabla_\theta w_{n,j}^{F,\theta,\alpha}}{w_{n,j}^{F,\theta,\alpha}}-\frac{1}{J}\sum_{j=1}^J \frac{\nabla_\theta w_{n,j}^{F,\theta,\alpha,k}}{w_{n,j}^{F,\theta,\alpha,k}}\right\rVert_2^2\\
    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta,\alpha}\right)-\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta,\alpha,k}\right)\right\rVert_2^2.
\end{align}

@@

    &  \left\lVert\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha}\right)-\nabla_\theta\log\left(\sum_{j=1}^J w_{n,j}^{F,\theta,\alpha,k}\right)\right\rVert_2^2\\
    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta,\alpha}\right)-\frac{1}{J}\sum_{j=1}^J \nabla_\theta \log\left(w_{n,j}^{F,\theta,\alpha,k}\right)\right\rVert_2^2 \\
    &= \left\lVert\frac{1}{J}\sum_{j=1}^J \nabla_\theta \left(\log\left(w_{n,j}^{F,\theta,\alpha}\right)-\log\left(w_{n,j}^{F,\theta,\alpha,k}\right)\right)\right\rVert_2^2\\

@@

We can then bound $\|\nabla_\theta\hat\ell^\alpha(\theta) - \nabla_\theta \hat s_k^\alpha(\theta)\|$ as follows:

@@

    &\left\lVert\nabla_\theta\hat\ell^\alpha(\theta) - \nabla_\theta \hat s_k^\alpha(\theta) \right\rVert_2^2

@@

    &= \left\lVert\sum_{n=1}^N \nabla_\theta \Bigg(\!\log\!\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha}\right) \!-\! \log\!\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha,k}\right)
    \!-\! \log\!\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,\alpha}\right) \!+\! \log \! \left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,\alpha,k}\right)\!\Bigg)\right\rVert_2^2 

@@

    &\leq \sum_{n=1}^N \left\lVert\nabla_\theta \left(\log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha}\right)- \log\left(\sum_{j=1}^Jw_{n,j}^{F,\theta,\alpha,k}\right)\right)\right\lVert_{\infty}

@@

    & \qquad\qquad +\sum_{n=1}^N \left\lVert\nabla_\theta \left(\log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,\alpha}\right) + \log\left(\sum_{j=1}^Jw_{n-1,j}^{A, F,\theta,\alpha,k}\right)\right)\right\rVert_2^2,

@@

\begin{equation}\left\lVert\nabla_\theta\hat\ell^\alpha(\theta) - \nabla_\theta \hat s_k^\alpha(\theta) \right\rVert_2^2 \leq  \frac{2\alpha^k}{1-\alpha}Np \, G'(\theta)^2,\end{equation}

@@

Now, we bound the variance of $\nabla_\theta \hat s_k^\alpha(\theta)$. 

@@

We can then decompose this expression with the derivative of the logarithm while noting that $w_{n,j}^{F,\theta,\alpha,k} = 1$ whenever $\theta=\phi$, to see that
